Compositional equivalences based on Open pNets

### 2022-03-24 2016-Weak-FH-Bisim/WeakBisim.tex
format jlamp + positioning again

\documentclass{lmcs}
%\overfullrule=1mm
\usepackage[multiple]{footmisc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
%\usepackage{setspace}
%\usepackage{indentfirst}
\usepackage{amsmath,amssymb,amscd,mathrsfs}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{epsfig,color,subfigure,enumitem,soul}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
\definecolor{darkgreen}{rgb}{0.1, 0.5, 0.1}
\newcommand{\LUDO}[1]{\textcolor{darkgreen}{#1}}
\newcommand{\RAB}[1]{\textcolor{magenta}{#1}}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\newcommand{\bigcupdot}{\charfusion[\mathop]{\bigcup}{\cdot}}
\makeatletter
\newcommand{\raisemath}[1]{\mathpalette{\raisem@th{#1}}}
\newcommand{\raisem@th}[3]{\raisebox{#1}{$#2#3$}}
\makeatother
\newcommand{\shortotimes}{\!\otimes\!}
\newcommand{\shortodot}{\!\odot\!}
\usepackage{macrospNets}

\newcommand{\nounderline}[1]{#1}

@@

\newtheorem{theorem}{Theorem}
%\newtheorem{prop}{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{algorithm}[theorem]{Algorithm}
%\newtheorem{remark}[theorem]{Remark}
\newtheorem{definition}{Definition}
\newtheorem{property}{Property}
\newtheorem{example}{Example}
%\newtheorem{problem}[theorem]{Problem}
%\newenvironment{proof}{\paragraph{Proof}}{}

@@

\title{Compositional equivalences based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\email{rabea.ameur-boulifa@telecom-paris.fr}

\author{Ludovic Henrio}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\email{ludovic.henrio@cnrs.fr}

\author{Eric Madelaine}
\address{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA, BP 93, 06902 Sophia Antipolis, France}
\email{eric.madelaine@inria.fr}
%\date{}                                           % Activate to display a given date or no date

@@

\subsubsection*{pNets and open automata}
pNet is a convenient model to model concurrent systems in a hierarchical and parameterised way. The coordination between processes is expressed as synchronisation vectors that allow the definition of complex and expressive synchronisation patterns.
Open pNets are pNets for which some elements in the hierarchy are still undefined, such undefined elements are called \emph{holes}; a hole can be \emph{filled} later by providing another pNet.
 The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no hole. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion. 
We thus define \emph{open automata} that can be seen as LTSs with parameters and holes. The transitions of open automata are much more complex than transitions of an LTS as the firing on a transition depends on parameters and actions that are symbolic. This article defines the notion of \emph{open transition} a transition that is symbolic in terms of parameters and coordinated actions.

Contrarily to pNets, open automata are not hierarchical structures, thus they are more convenient for formal reasoning but not adapted to the definition of a complex and structured system like pNets. Additionally, open transitions are expressed in terms of logics more than in terms of synchronised actions, synchronisation vectors on the contrary make it easy to express synchronisations that exist in process algebra or in specification and high-level languages for distributed systems.

This article defines pNets and illustrates with an example how they can be used to provide the model of a communicating system. Then we introduce open automata to provide a semantics to open pNets and introduce bisimulation relations and their properties.
Open automata can be also seen as an algebra that can be studied independently from its application to pNets but their composition properties make more sense in a hierarchical model like pNets.

@@

We define two kinds of parameterised actions. The first kind distinguishes input variables and the second kind does not. 
We first define the set of actions that distinguish input variables, they will be used in the definition of \pLTS\ below:
\[
\begin{array}[l]{rcl@{\quad}p{7.5cm}}
  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
  p_i&::=& ?x~|~e_i&\text{parameters (input variable or expression)}\\
  e_i&::=& \symb{Value}~|~x~|~\symb{op}(\symb{e}_1,..,\symb{e}_n)&\text{Expressions}
\end{array}
\]
The \emph{input variables} in an action term are those marked with a
$\symb{?}$.
We additionally impose that each input variable does not
appear somewhere else in the same action term:
$p_i=?x\Rightarrow\forall j\neq i.\, x\notin \vars(p_j)$.
We define $iv(t)$  as the set of input variables of a term $t$ (without the '?' marker).
Action algebras can encode naturally usual point-to-point message passing calculi (using 
$a(?x_1,...,?x_n)$ for inputs, $a(v_1,..,v_n)$ for outputs), but it also allows
for more general synchronisation mechanisms, like gate negotiation in Lotos, or broadcast
communications.

@@

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{When using this notation, we suppose, without loss of generality that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following. Thus,
 $\subst{\Post}$ is  the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\odot$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortodot\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\

\[(x_{k}\gets e_{k})^{k\in K}\shortodot (x'_{k'}\gets e'_{k'})^{k'\in K'} =  
(x_{k}\gets e_{k}\subst{(x'_{k'}\gets e'_{k'})^{k'\in K'}})^{k\in K} \cup (x'_{k'}\gets e'_{k'})^{k'\in K''}\]
where $K''=\{k'\in K'|x'_{k'}\not\in\{x_k\}^{k\in K}\}$
%
%\LUDO{En fait comme on supose qu il y a pas de variable globale et donc que les post doivent affecter toutes les variables ce contre-exemple n est pas applicable! apres on peut definir comme ca si on veut ceinture et bretelles mais ca alourdit un peu la definitions}
%
%\ERIC{Heu Non... 
%  En pratique, dans la composition de posts de la regle WT3, les posts assignent bien toutes les variables de leur OT (strong), mais dans la composition, on fini par avoir des trous... Voir page 21-22. On est tombes tous les deux separemment, Rabea et moi, sur ce genre de cas.}
% 
%\LUDO{euuh ca doit etre lie qu fait qu il manque des ec leftarrowec dans les selfloop de l'OA de spec. Plutot que changer la def et les preuves je propose de rajouter la condition $vars(s')=dom(Post)$ pour tous les OAs ce qui ne change rien a part pour les selfloop que l on s autorisait apparemment a ecrire en omettant $x:=x$ ... au pire l'identite est un post toujours correct et valide}

@@

\begin{definition}[pNet composition]
	An open pNet: $\pNet = \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\,\myrangle$
 can be (partially) filled by providing  a pNet $\pNetQ$ to fill one of  its holes.	
	Suppose $j_0\in J$ and $\Sortop(\pNetQ) \subseteq \Sort_{j_0}$, then: % are compatible:
	%% \[\mylangle \pNet_i^{i\in I},S_j^{j\in J}, \symb{SV}_k^{k\in
	%%   K}\myrangle\left[(\pNet'_i)^{i\in L}\right]= \mylangle \pNet_i^{i\in
	%%   I}\left[(\pNet'_i)^{i\in L}\right]\uplus(\pNet'_i)^{i\in J\cap L},S_j^{j\in 
	%%J\setminus L},
	%% \symb{SV}_k^{k\in K}\myrangle
	%% \]
	\[\pNet\left[\pNetQ\right]_{j_0}= \mylangle 
	\pNet_i^{i\in I}\uplus\{j_0\mapsto \pNetQ\},\Sort_j^{j\in J\setminus \{j_0\}},
	\set{\symb{SV}}\,\myrangle
	\]
\end{definition}

@@

Figure \ref{SimpleProt:Impl} shows the pNet model of the protocol implementation, called $\symb{SimpleProtocolImpl}$. When the $\symb{Medium}$ detects an error (modelled by a local $\tau$ action), it sends back a $\texttt{m-error}$ message to the $\symb{Sender}$. The $\symb{Sender}$ increments its local error counter $ec$, and resends the message (including $ec$) to the $\symb{Medium}$, that will, eventually, transmit $m,ec$ to the $\symb{Receiver}$. 
%Strictly speaking, as variables are local to each  state, we have two distinct $ec$ variables in the \textit{Sender} pLTS, one in state \texttt{s1}, another one in state \texttt{s2}, the value of the variable is updated along the transitions entering each of the two states.

@@

\item	The second rule deals with pNet nodes: for each possible
	synchronisation vector (of index $k$) applicable to the rule subject, the premisses
	include one {\em open transition} for each sub-pNet involved, one possible
	{\em action} for each hole involved, and the predicate relating these
	with the resulting action of the vector. The sub-pNets involved are split between two 
	sets, $I_2$ for sub-pNets that are pLTSs (with open transitions obtained by rule \textbf{Tr1}), and $I_1$ for the sub-pNets that are not pLTSs (with open transitions obtained by rule \textbf{Tr2}), $J$ is the set of 
	holes involved in the transition\footnote{Formally, if $SV_k \!=\! \SV{({\alpha'})_m^{m 
	\in M}}{\alpha'}{e_b}$ is a synchronisation vector  of \pNet\  then $J=M\cap 
	\Holes(\pNet)$, $I_2=M\cap \Leaves(\pNet)$,  $I_1=M\setminus J \setminus 
	I_2$}\footnote{We could replace $I_1$ and $I_2$ by their formal definition in \textbf{Tr2} but the rule would be more difficult to read.}.                                                                    
\begin{mathpar}
    \mprset {vskip=.45ex}
\inferrule
    {
\Leaves(\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{\,k\in 
    	K}\myrangle) \!=\! \pLTS_l^{\,l\in L} \quad  	
k\!\in\! K \quad SV_k \!=\! \SV{(\alpha'_m)^{m \in I_1\uplus I_2\uplus J}}{\alpha'}{e_b} 
\\
\\     	
	\forall m\!\!\in\!\! I_1. {\pNet_m 
	\models\openrule
    	{\beta_{j}^{j\in J_m}, \Pred_m, \Post_m}
    	{\ostate{s_{i}^{i \in L_m}} \OTarrow {\alpha_m}
    		\ostate{(s_i^\prime)^{i\in L_m}}} }	
  \qquad
\forall m\!\!\in\!\! I_2.		{ \pNet_m 
    	 \models
    	\openrule
    	{\emptyset, \Pred_m, \Post_m}
    	{\ostate{s_m} \OTarrow {\alpha_m}
    		\ostate{s_m'}} }\\\\
    J' = \biguplus_{m\in I_1}\!\! J_m \uplus J	\\
    	\Pred = \bigwedge_{m\in I_1\uplus I_2}\!\! \Pred_m \land
    	\Predsv(SV_k,\alpha_m^{m\in I_1\uplus I_2},\beta_j^{j\in J},\alpha)\\ 
    	\forall i\in	L\backslash \left(\biguplus_{m\in I_1}\!\! L_m \uplus I_2\right).\,s'_i=s_i \\
    \fresh(\alpha'_m,\alpha',\beta_j^{j\in J},\alpha) 
    }
    {\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{\,k\in K}\myrangle
    	\models
    	{\openrule
    		{
    		\beta_j^{j\in J^\prime}, \Pred,  \biguplus_{m\in I_1\uplus I_2} 
    		\Post_m}
    		{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
    			\ostate{(s_i^\prime)^{i\in L}}}
    	}
    }\quad {\TrDeux}
\end{mathpar} 
\end{itemize} 
	\medskip
%        \TODO{may be explain how $\Pred(SV,a_i^{i\in I_k},b_j^{j\in
%            J_k},v)$ is built ? You mean more than what is written on previous page????}
	%%    \TODO{I have tentatively added the sort constraint on hole actions, that was
	%%not included in the first version... I'm unsure whether this is the best place to
	%%include it, because it may change the decidability conditions on predicates}
\end{definition}
        	A key to understand this rule is that the open transitions are
	expressed in terms of the leaves and holes of the whole pNet structure,
	i.e. a flatten view of the pNet. For example, $L$ is the index set of the
	Leaves, $L_m$ the index set of the leaves of one sub-pNet indexed $m$, so all $L_m$
	are disjoint subsets of $L$. Thus the states in the open transitions,
	at each level, are tuples including states of all the
	leaves of the pNet, not only those involved in the chosen
	synchronisation vector.

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/SPSpecOpen}}
   \caption{Open automaton for  $\symb{SimpleProtocolSpec}$}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

@@

 \begin{figure}[ht]
  \centerline{\includegraphics[width=15cm]{XFIG/SPImplOpen}}
  \caption{Open automaton for $\symb{SimpleProtocolImpl}$}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=13cm]{XFIG/WOT2-result}}
  \caption{Construction of an example of weak open transition}
   \label{WOT2}
\end{figure}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=15cm]{XFIG/SPSpecWeakOpen}}
  \caption{Weak Open Automaton of  $\symb{SimpleProtocolSpec}$}
   \label{SimpleProtCounter:WeakSpecOA}
\end{figure}

@@

$ \forall n\ge 0.\, WI_3(n) = \openrule
  {\{\texttt{P}\mapsto \texttt{p-send(m)}\}, True,
    (\texttt{s\_msg}\gets \texttt{m}, \texttt{s\_ec}\gets n)}
  {000 \OTWeakarrow {\nounderline{\texttt{in(m)}}} 100}
        $

$  \forall n\ge 0.\, WI_4(n) = \openrule
         {\{\}, True, 
   (\texttt{m\_msg}\!\gets\! \texttt{s\_msg}, \texttt{m\_ec}\!\gets\! \texttt{s\_ec}+n, \texttt{s\_ec}\!\gets\! \texttt{s\_ec}+n)}
         {100 \OTWeakarrow {\tau} 210}
        $

$  \forall n\ge 0.\, WI_{456}(n) = \openrule
         {\{\}, True, 
    (\texttt{s\_ec}\gets \texttt{s\_ec}+n)}
  {100 \OTWeakarrow {\tau} 100}
        $

>>>>

\documentclass{elsarticle}
%\overfullrule=1mm
\usepackage[multiple]{footmisc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{geometry}                % See geometry.pdf to learn the layout
%\geometry{a4paper}                   % ... or a4paper or a5paper or ...
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{DotArrow}
\usepackage{hyperref}
\usepackage{listings}
\lstset{basicstyle=\small}

%\usepackage{amsfonts}
%\usepackage{fancyhdr}
%\usepackage{cite}
%\usepackage{ifthen}
%\usepackage{amssymb}
%\usepackage{fancyhdr}
%\usepackage{pifont}
\usepackage{stmaryrd}
\usepackage{mathtools,mathpartir}
\usepackage{proof}
\usepackage{makecell}
%\usepackage{indentfirst}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usepackage{epsfig,color,subfigure,enumitem,soul}
\newcommand{\TODO}[1]{\textcolor{red}{\textbf{[TODO:#1]}}}
\newcommand{\NOTE}[1]{\textcolor{blue}{\textbf{[NOTE:#1]}}}
\newcommand{\ERIC}[1]{\textcolor{blue}{#1}}
\definecolor{darkgreen}{rgb}{0.1, 0.5, 0.1}
\newcommand{\LUDO}[1]{\textcolor{darkgreen}{#1}}
\newcommand{\RAB}[1]{\textcolor{magenta}{#1}}
\newcommand{\coloncolon}{{:\hspace{-.2ex}:}}
\newcommand{\bigcupdot}{\charfusion[\mathop]{\bigcup}{\cdot}}
\makeatletter
\newcommand{\raisemath}[1]{\mathpalette{\raisem@th{#1}}}
\newcommand{\raisem@th}[3]{\raisebox{#1}{$#2#3$}}
\makeatother
\newcommand{\shortotimes}{\!\otimes\!}
\newcommand{\shortodot}{\!\odot\!}
\usepackage{macrospNets}
\usepackage{amsmath,amssymb,amscd,mathrsfs}
\newcommand{\nounderline}[1]{#1}

@@

\newtheorem{thm}{Theorem}
%\newtheorem{prop}{Proposition}
%\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{lem}{Lemma}
\newtheorem{algorithm}[thm]{Algorithm}
%\newtheorem{remark}[theorem]{Remark}
\newdefinition{definition}{Definition}
\newtheorem{property}{Property}
\newdefinition{example}{Example}
%\newtheorem{problem}[theorem]{Problem}
\newproof{proof}{Proof}

@@

\title{Compositional equivalences based on Open pNets
%\thanks{This work was partially funded by the Associated Team FM4CPS
%  between INRIA and ECNU, Shanghai}
}
\author{Rab\'ea Ameur-Boulifa}
\address{LTCI, T\'el\'ecom Paris, Institut Polytechnique de Paris, France
}         
\ead{rabea.ameur-boulifa@telecom-paris.fr}

\author{Ludovic Henrio}
\address{Univ Lyon, EnsL, UCBL, CNRS, Inria,  LIP, F-69342, LYON Cedex 07, France.}
\ead{ludovic.henrio@cnrs.fr}

\author{Eric Madelaine}
\address{INRIA Sophia Antipolis M\'edit\'erann\'ee, UCA, BP 93, 06902 Sophia Antipolis, France}
\ead{eric.madelaine@inria.fr}
%\date{}                                           % Activate to display a given date or no date

@@

\subsubsection*{pNets}
pNet is a convenient model to model concurrent systems in a hierarchical and parameterised way. The coordination between processes is expressed as synchronisation vectors that allow the definition of complex and expressive synchronisation patterns.
Open pNets are pNets for which some elements in the hierarchy are still undefined, such undefined elements are called \emph{holes}; a hole can be \emph{filled} later by providing another pNet.
This article first defines pNets and illustrates with an example how they can be used to provide the model of a communicating system.

\subsubsection{A semantics for Open pNets based on open automata}
The semantics of pNets can be expressed as a translation to a labelled transition system, but only if the pNet has no parameter and no hole. Adding parameters to a LTS is quite standard but enabling holes inside LTSs is not a well-defined notion.

To define a semantics for open pNets we thus need LTSs that can have two kinds of parameters: \emph{variable} and \emph{processes}. We call such LTSs with parameters and holes and equiped with a composition operation, \emph{open automata}. The main goal of this article is to define the theory behind open automata and to use them to provide a semantics and prove composition properties for open pNets.
  The transitions of open automata are much more complex than transitions of an LTS as the firing on a transition depends on parameters and actions that are symbolic. This article defines the notion of \emph{open transition} a transition that is symbolic in terms of parameters and coordinated actions.

Contrarily to pNets, open automata are not hierarchical structures, we consider them here as a mathematical structure convenient for formal reasoning but not adapted to the definition of a complex and structured system. 
Open transitions are expressed in terms of logics while the communication in pNets is specified as synchronisation vectors that specify synchronised actions.
Open automata could alternatively be  seen as an algebra that can be studied independently from its application to pNets but their composition properties make more sense in a hierarchical model like pNets.

@@

We define two kinds of parameterised actions. The first kind distinguishes input variables and the second kind does not. 
We first define the set of actions that distinguish input variables, they will be used in the definition of \pLTS\ below:
\[
\begin{array}[l]{rcl@{\quad}p{5.65cm}}
  \alpha\in\AlgA&::=&a(p_1,\ldots,p_n)&\text{action terms}\\
  p_i&::=& ?x~|~e_i&\text{parameters (input var or expression)}\\
  e_i&::=& \symb{Value}~|~x~|~\symb{op}(\symb{e}_1,..,\symb{e}_n)&\text{Expressions}
\end{array}
\]
The \emph{input variables} in an action term are those marked with a
$\symb{?}$.
We additionally impose that each input variable does not
appear somewhere else in the same action term:
$p_i=?x\Rightarrow\forall j\neq i.\, x\notin \vars(p_j)$.
We define $iv(t)$  as the set of input variables of a term $t$ (without the '?' marker).
Action algebras can encode naturally usual point-to-point message passing calculi (using 
$a(?x_1,...,?x_n)$ for inputs, $a(v_1,..,v_n)$ for outputs), but it also allows
for more general synchronisation mechanisms, like gate negotiation in Lotos, or broadcast
communications.

@@

This article also uses substitutions. Applying a substitution inside a term $t$ is denoted $t\subst{y_i\gets e_i}^{i\in I}$ and consists in replacing in parallel all the occurrences of variables $y_i$ in the term $t$ by the terms $e_i$. Note that a substitution is defined by a partial function that is applied on the variables inside a term. We let $\Post$ range over partial functions that are used as substitution and use the notation $\{y_i\gets e_i\}^{i\in I}$ to define such a partial function\footnote{When using this notation, we suppose, without loss of generality that each $y_i$ is different.}. These partial functions are sometimes called \emph{substitution functions} in the following. Thus,
 $\subst{\Post}$ is  the operation that applies,  in a parallel manner,  the substitution defined by the partial function $\Post$. $\odot$ is a composition operator on these partial functions, such that for any term $t$ we have: $t\subst{\Post\shortodot\Post'} = (t\subst{\Post'})\subst{\Post}$.
This property must also be valid when the substitution does not operate on all variables.
We thus define a composition operation as follows: 
%\TODO{version d eric reprise et corrigee}
%$(x_k\gets e_k)^{k\in K}\shortotimes \Post =  (x_k\gets e_k\subst{\Post})^{k\in K}$
%\ERIC{J'ai un probleme ici... quand $Post'$ agit sur des vars de $t$ qui ne sont pas substitutees par $Post$, alors les 2 defs de $\shortotimes$ ci-dessus ne sont pas equivalentes:}\\
%$F(x,z)(\subst{x\gets z}\shortotimes\subst{z\gets 1})=F(x,z)(\subst{x\gets 1})=F(1,z)$\\
%\ERIC{Alors que}\\
%$(F(x,z)\subst{x\gets z})\subst{z\gets 1})=F(z,z)\subst{z\gets 1}=F(1,1)$
%
%\ERIC{Une definition plus correcte serait peut-etre:}\\
\begin{align*}
(x_{k}\!\gets\! e_{k})^{k\in K}\odot (x'_{k'}\gets e'_{k'})^{k'\in K'} \!=&  
(x_{k}\gets e_{k}\subst{(x'_{k'}\!\gets\! e'_{k'})^{k'\in K'}})^{k\in K} \\&\cup (x'_{k'}\!\gets\! e'_{k'})^{k'\in K''}
\end{align*}where $K''=\{k'\in K'|x'_{k'}\not\in\{x_k\}^{k\in K}\}$
%
%\LUDO{En fait comme on supose qu il y a pas de variable globale et donc que les post doivent affecter toutes les variables ce contre-exemple n est pas applicable! apres on peut definir comme ca si on veut ceinture et bretelles mais ca alourdit un peu la definitions}
%
%\ERIC{Heu Non... 
%  En pratique, dans la composition de posts de la regle WT3, les posts assignent bien toutes les variables de leur OT (strong), mais dans la composition, on fini par avoir des trous... Voir page 21-22. On est tombes tous les deux separemment, Rabea et moi, sur ce genre de cas.}
% 
%\LUDO{euuh ca doit etre lie qu fait qu il manque des ec leftarrowec dans les selfloop de l'OA de spec. Plutot que changer la def et les preuves je propose de rajouter la condition $vars(s')=dom(Post)$ pour tous les OAs ce qui ne change rien a part pour les selfloop que l on s autorisait apparemment a ecrire en omettant $x:=x$ ... au pire l'identite est un post toujours correct et valide}

@@

\begin{definition}[pNet composition]
	An open pNet: $\pNet \!=\! \mylangle \pNet_i^{i\in I}, \Sort_j^{j\in J}, 
	\set{\symb{SV}}\,\myrangle$
 can be (partially) filled by providing  a pNet $\pNetQ$ to fill one of  its holes.	
	Suppose $j_0\in J$ and $\Sortop(\pNetQ) \subseteq \Sort_{j_0}$, then: % are compatible:
	%% \[\mylangle \pNet_i^{i\in I},S_j^{j\in J}, \symb{SV}_k^{k\in
	%%   K}\myrangle\left[(\pNet'_i)^{i\in L}\right]= \mylangle \pNet_i^{i\in
	%%   I}\left[(\pNet'_i)^{i\in L}\right]\uplus(\pNet'_i)^{i\in J\cap L},S_j^{j\in 
	%%J\setminus L},
	%% \symb{SV}_k^{k\in K}\myrangle
	%% \]
	\[\pNet\left[\pNetQ\right]_{j_0}= \mylangle 
	\pNet_i^{i\in I}\uplus\{j_0\mapsto \pNetQ\},\Sort_j^{j\in J\setminus \{j_0\}},
	\set{\symb{SV}}\,\myrangle
	\]
\end{definition}

@@

\fussy
Figure \ref{SimpleProt:Impl} shows the pNet model of the protocol implementation, called $\symb{SimpleProtocolImpl}$. When the $\symb{Medium}$ detects an error (modelled by a local $\tau$ action), it sends back a $\texttt{m-error}$ message to the $\symb{Sender}$. The $\symb{Sender}$ increments its local error counter $ec$, and resends the message (including $ec$) to the $\symb{Medium}$, that will, eventually, transmit $m,ec$ to the $\symb{Receiver}$. 
\sloppy
%Strictly speaking, as variables are local to each  state, we have two distinct $ec$ variables in the \textit{Sender} pLTS, one in state \texttt{s1}, another one in state \texttt{s2}, the value of the variable is updated along the transitions entering each of the two states.

@@

\item	The second rule deals with pNet nodes: for each possible
	synchronisation vector (of index $k$) applicable to the rule subject, the premisses
	include one {\em open transition} for each sub-pNet involved, one possible
	{\em action} for each hole involved, and the predicate relating these
	with the resulting action of the vector. The sub-pNets involved are split between two 
	sets, $I_2$ for sub-pNets that are pLTSs (with open transitions obtained by rule \textbf{Tr1}), and $I_1$ for the sub-pNets that are not pLTSs (with open transitions obtained by rule \textbf{Tr2}), $J$ is the set of 
	holes involved in the transition\footnote{Formally, if $SV_k \!=\! \SV{({\alpha'})_m^{m 
	\in M}}{\alpha'}{e_b}$ is a synchronisation vector  of \pNet\  then $J=M\cap 
	\Holes(\pNet)$, $I_2=M\cap \Leaves(\pNet)$,  $I_1=M\setminus J \setminus 
	I_2$}\footnote{We could replace $I_1$ and $I_2$ by their formal definition in \textbf{Tr2} but the rule would be more difficult to read.}.   
\end{itemize}                                                                  
\begin{mathpar}
    \mprset {vskip=.45ex}
\inferrule
    {
\Leaves(\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{\,k\in 
    	K}\myrangle) \!=\! \pLTS_l^{\,l\in L} \quad  	
k\!\in\! K \\\\ SV_k \!=\! \SV{(\alpha'_m)^{m \in I_1\uplus I_2\uplus J}}{\alpha'}{e_b} 
\\
\\     	
	\forall m\!\!\in\!\! I_1. {\pNet_m 
	\models\openrule
    	{\beta_{j}^{j\in J_m}, \Pred_m, \Post_m}
    	{\ostate{s_{i}^{i \in L_m}} \OTarrow {\alpha_m}
    		\ostate{(s_i^\prime)^{i\in L_m}}} }	
  \\\\
\forall m\!\!\in\!\! I_2.		{ \pNet_m 
    	 \models
    	\openrule
    	{\emptyset, \Pred_m, \Post_m}
    	{\ostate{s_m} \OTarrow {\alpha_m}
    		\ostate{s_m'}} }\qquad
    J' = \biguplus_{m\in I_1}\!\! J_m \uplus J	\\
    	\Pred = \!\!\bigwedge_{m\in I_1\uplus I_2}\!\! \Pred_m \land
    	\Predsv(SV_k,\alpha_m^{m\in I_1\uplus I_2},\beta_j^{j\in J},\alpha)\\ 
    	\forall i\in	L\backslash \left(\biguplus_{m\in I_1}\!\! L_m \uplus I_2\right).\,s'_i=s_i \\
    \fresh(\alpha'_m,\alpha',\beta_j^{j\in J},\alpha) 
    }
    {\mylangle {\pNet}_m^{m\in I}, \set{\Sort}, \symb{SV}_k^{\,k\in K}\myrangle
    	\models
    	{\openrule
    		{
    		\beta_j^{j\in J^\prime}, \Pred,  \biguplus_{m\in I_1\uplus I_2} 
    		\Post_m}
    		{\ostate{s_i^{i\in L}} \OTarrow {\alpha}
    			\ostate{(s_i^\prime)^{i\in L}}}
    	}
    }\quad {\TrDeux}
\end{mathpar}

	\medskip
%        \TODO{may be explain how $\Pred(SV,a_i^{i\in I_k},b_j^{j\in
%            J_k},v)$ is built ? You mean more than what is written on previous page????}
	%%    \TODO{I have tentatively added the sort constraint on hole actions, that was
	%%not included in the first version... I'm unsure whether this is the best place to
	%%include it, because it may change the decidability conditions on predicates}
\end{definition}
        	A key to understand this rule is that the open transitions are
	expressed in terms of the leaves and holes of the whole pNet structure,
	i.e. a flatten view of the pNet. For example, $L$ is the index set of the
	Leaves, $L_m$ the index set of the leaves of one sub-pNet indexed $m$, so all $L_m$
	are disjoint subsets of $L$. Thus the states in the open transitions,
	at each level, are tuples including states of all the
	leaves of the pNet, not only those involved in the chosen
	synchronisation vector.

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=\linewidth]{XFIG/SPSpecOpen}}
   \caption{Open automaton for  $\symb{SimpleProtocolSpec}$}
   \label{SimpleProtCounter:SpecOA}
\end{figure}

@@

 \begin{figure}[ht]
  \centerline{\includegraphics[width=\linewidth]{XFIG/SPImplOpen}}
  \caption{Open automaton for $\symb{SimpleProtocolImpl}$}  \label{SimpleProtCounter:ImplOA}
\end{figure}

@@

\begin{figure}[ht]
   \centerline{\includegraphics[width=\linewidth]{XFIG/WOT2-result}}
  \caption{Construction of an example of weak open transition}
   \label{WOT2}
\end{figure}

@@

\begin{figure}[h]
   \centerline{\includegraphics[width=\linewidth]{XFIG/SPSpecWeakOpen}}
  \caption{Weak Open Automaton of  $\symb{SimpleProtocolSpec}$}
   \label{SimpleProtCounter:WeakSpecOA}
\end{figure}

@@

$ \forall n\ge 0.\, \symb{WI}_3(n) = \openrule
  {\{\texttt{P}\mapsto \texttt{p-send(m)}\}, \symb{True},
    (\texttt{s\_msg}\gets \texttt{m}, \texttt{s\_ec}\gets n)}
  {000 \OTWeakarrow {\nounderline{\texttt{in(m)}}} 100}
        $

$\hspace{-6pt}  \forall n\!\ge\! 0.\, \symb{WI}_4(n) \!=\! \openrule
         {\{\}, \symb{True}, 
   (\texttt{m\_msg}\!\gets\! \texttt{s\_msg}, \texttt{m\_ec}\!\gets\! \texttt{s\_ec}\!+\!n, \texttt{s\_ec}\!\gets\! \texttt{s\_ec}\!+\!n)}
         {100 \OTWeakarrow {\tau} 210}
        $

$  \forall n\ge 0.\, \symb{WI}_{456}(n) = \openrule
         {\{\}, \symb{True}, 
    (\texttt{s\_ec}\gets \texttt{s\_ec}+n)}
  {100 \OTWeakarrow {\tau} 100}
        $
